Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-10-24 paper-new/concrete-relational-model.tex
changes we worked on together

\begin{remark}
  The reason why the set of syntactic perturbations needs to carry a monoid
  structure and why the interpretation needs to be a homomorphism of monoids is
  related to the construction of the identity relations and composition of
  relations on value/computation types. As part of this construction, we will
  need to produce a syntactic perturbation corresponding to the identity
  predomain/error domain relation, as well as corresponding to the composition
  of relations. The monoid structure gives us what we need to produce these
  perturbations.
  % we will need to produce a syntactic perturbation corresponding to the identity
  % relations $r(A)$, as well as one for the composition of relations. The monoid
  % structure gives us what we need to produce these elements. Likewise, the
  % reason that the interpretations as semantic perturbations must be
  % \emph{homomorphisms} of monoids is that we need a syntactic perturbation that
  % is interpreted as the identity semantic perturbation, and likewise given two
  % syntactic perturbations $\delta_1$ and $\delta_2$ we need a syntactic
  % pertubration $\delta_{1,2}$ that is interpreted as the composition of the
  % semantic perturbations $i_A(\delta_1)$ and $i_A(\delta_2)$.
  % These properties are used to define the identity and composition of value and computation relations
\end{remark}

% It doesn't work: not all types have such an action, e.g., the function type, because of the definition of the perturbations
% The fact that we cannot define the morphisms this way is unfortunate, since doing so would simplify
% some aspects of the model construction.
\begin{remark}
  It might seem strange that in defining the morphisms of value and computation
  objects, there is no requirement involving the monoids. For example, one could
  imagine defining a morphism between value objects $(A, M_A, i_A)$ and $(A',
  M_{A'}, i_{A'})$ to be a predomain morphism $f \colon A \to A'$ along with a
  monoid homomorphism $h \colon M_A \to M_{A'}$ such that $f$ and $h$ satisfy a
  compatibility condition with respect to the interpretations $i_A$ and
  $i_{A'}$. However, with this definition we cannot interpret the function type,
  since the morphisms defined in this way do not themselves form a value type. 
  % This is due to the way that the syntactic perturbations are defined for the function type.
  %
  The fact that we cannot define the morphisms in this way is unfortunate, since
  doing so would simplify some aspects of the model construction.

@@

Finally, the most interesting type is of course the dynamic type. 
It may seem as though we need to define the perturbations for $D$ by
guarded recursion, but in fact we define them via least-fixpoint
in the category of monoids:
%
\( M_D \cong (M_{D \times D}) \oplus M_{U(D \to \li D)}. \)
%
\begin{remark}
  The reason for defining the perturbations inductively is technical and has to
  do with the fact that the perturbations corresponding to the predomain $\later
  A$ need to be the same as the perturbations of $A$, \emph{not} later of the
  perturbations of $A$. There may be a way to make this work by changing some
  definitions; we leave this to future investigation.
\end{remark}
%
We now explain how to interpret these perturbations as endomorphisms.
We define $i_D : M_D \to \{ f : D \to D \mid f \bisim \id \}$ by induction as follows:
%
\begin{align*}
  i_D(m_\times) &= \lambda d. \text{case $d$ of }
    \{ \itimes (d_1, d_2) \To \itimes (i_{D \times D}(m_\times)(d_1, d_2)) \alt x \To x \} \\
  i_D(m_\to) &= \lambda d. \text{case $d$ of }
    \{ \iarr (\tilde{f}) \To \iarr (\lambda t. i_{U(D \to \li D)}(m_\to)(\tilde{f}_t)) \alt x \To x \}
\end{align*}
%
In the case of a perturbation on $D \times D$, we use the interpretation for the
perturbations of $D \times D$ as defined earlier in this section, which in turn
will use $i_D$ inductively. Similarly, for a perturbation on $U(D \to \li D)$ we
use the interpretation for perturbations on $U(D \to \li D)$.
One can verify that this defines a homomorphism from $M_D \to \{ f : D \to D : f
\bisim \id \}$.

>>>>

\begin{remark}
  Syntactic perturbations are equipped with a monoid structure in
  order to support identity and composition relations on value and
  computation types. That is, the identity relation requires an
  identity perturbation and composition of relations requires a
  composition operation on perturbations.
  %% The reason why the set of syntactic perturbations needs to carry a monoid
  %% structure and why the interpretation needs to be a homomorphism of monoids is
  %% related to the construction of the identity relations and composition of
  %% relations on value/computation types. As part of this construction, we will
  %% need to produce a syntactic perturbation corresponding to the identity
  %% predomain/error domain relation, as well as corresponding to the composition
  %% of relations. The monoid structure gives us what we need to produce these
  %% perturbations.
  % we will need to produce a syntactic perturbation corresponding to the identity
  % relations $r(A)$, as well as one for the composition of relations. The monoid
  % structure gives us what we need to produce these elements. Likewise, the
  % reason that the interpretations as semantic perturbations must be
  % \emph{homomorphisms} of monoids is that we need a syntactic perturbation that
  % is interpreted as the identity semantic perturbation, and likewise given two
  % syntactic perturbations $\delta_1$ and $\delta_2$ we need a syntactic
  % pertubration $\delta_{1,2}$ that is interpreted as the composition of the
  % semantic perturbations $i_A(\delta_1)$ and $i_A(\delta_2)$.
  % These properties are used to define the identity and composition of value and computation relations
\end{remark}

% It doesn't work: not all types have such an action, e.g., the function type, because of the definition of the perturbations
% The fact that we cannot define the morphisms this way is unfortunate, since doing so would simplify
% some aspects of the model construction.
\begin{remark}
  It might seem strange that we equip objects with a monoid of
  perturbations but we do not require morphisms to respect this
  structure in any way. A natural attempt would be defining a morphism
  between value objects $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ to
  be a predomain morphism $f \colon A \to A'$ along with a monoid
  homomorphism $h \colon M_A \to M_{A'}$ such that $f$ and $h$ satisfy
  a compatibility condition with respect to the interpretations $i_A$
  and $i_{A'}$. However, this condition simply fails to hold for all
  terms, in particular the property is preserved by $\lambda$
  abstraction.
  %% with this definition we cannot interpret the
  %% function type, since the morphisms defined in this way do not
  %% themselves form a value type.
  % This is due to the way that the syntactic perturbations are defined for the function type.
  %
  %% The fact that we cannot define the morphisms in this way is unfortunate, since
  %% doing so would simplify some aspects of the model construction.

@@

Finally, the most interesting type is of course the dynamic type. 
It may seem as though we need to define the perturbations for $D$ by
guarded recursion, but in fact we define them via least-fixpoint
in the category of monoids:
%
\( M_D \cong (M_{D \times D}) \oplus M_{U(D \to \li D)}. \)
%
\begin{remark}
  Here it is crucial to define the perturbations of the dynamic type
  \emph{inductively} rather than by guarded recursion. If we were to
  define the perturbations by guarded recursion, then it would be
  impossible to define the pull homomorphism for the $\iarr$ relation.
\end{remark}
%
We now explain how to interpret these perturbations as endomorphisms.
We define $i_D : M_D \to \{ f : D \to D \mid f \bisim \id \}$ by induction as follows:
%
\begin{align*}
  i_D(m_\times) &= \lambda d. \text{case $d$ of }
    \{ \itimes (d_1, d_2) \To \itimes (i_{D \times D}(m_\times)(d_1, d_2)) \alt x \To x \} \\
  i_D(m_\to) &= \lambda d. \text{case $d$ of }
    \{ \iarr (\tilde{f}) \To \iarr (\lambda t. i_{U(D \to \li D)}(m_\to)(\tilde{f}_t)) \alt x \To x \}
\end{align*}
%
In the case of a perturbation on $D \times D$, we use the interpretation for the
perturbations of $D \times D$ as defined earlier in this section, which in turn
will use $i_D$ inductively. Similarly, for a perturbation on $U(D \to \li D)$ we
use the interpretation for perturbations on $U(D \to \li D)$.
One can verify that this defines a homomorphism from $M_D \to \{ f : D \to D : f
\bisim \id \}$.

### 2024-10-24 paper-new/intro.tex
changes we worked on together

\subsection{Synthetic Guarded Domain Theory}

@@

% The move to synthetic setting is useful, but now you're stuck in the guarded setting
The simplicity afforded by working in the synthetic setting comes with a
drawback: by changing our ambient foundations so that all types represent
step-indexed sets, it is no longer possible to define a type that denotes an
ordinary set. Roughly speaking, what is needed is an ``escape hatch'' to leave
the guarded setting and make definitions that are interpreted as constructions
in ordinary mathematics. This is typically accomplished by adding a modality to
the guarded type theory. The specific approach we take in this paper is the
technique of \emph{clock-quantification} \cite{atkey-mcbride2013}. In
particular, the guarded type theory in which we work has a notion of clocks that
index all of our guarded constructions. This is discussed in more detail in
Section \ref{sec:big-step-term-semantics}.

@@

>>>>

@@

@@

% The move to synthetic setting is useful, but now you're stuck in the guarded setting
We have to take care when defining such a big-step semantics in
guarded type theory, as a naive definition of termination is
incorrect: we can prove using guarded recursion that an infinite loop
terminates! This is because the definition of termination inside
guarded logic has a different meaning, essentially meaning that for
every finite number $n$, either the program terminates in $< n$ steps
or it takes $n+1$ steps, which is equivalent to a statement of
progress rather than termination. For this reason, we need a kind of
``escape hatch'' in our ambient type theory to leave the guarded
setting and make definitions that are interpreted as constructions in
ordinary mathematics. We incorporate this by using
\emph{clock-quantification}\max{Eric should provide a citation for Clocked cubical here since that's close to our actual approach} \cite{atkey-mcbride2013} in our type
theory. In particular, the guarded type theory in which we work has a
notion of clocks that index all of our guarded constructions. This is
discussed in more detail in Section \ref{sec:big-step-term-semantics}.
